\documentclass[12pt,a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{array}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage[style=apa, backend=biber]{biblatex}
\addbibresource{references.bib}
\usepackage{hyperref}
\usepackage{makecell}
\usepackage{booktabs} % For professional looking tables
\usepackage{tikz}
\usetikzlibrary{
    positioning,
    arrows.meta,
    shapes.geometric,
    shapes.misc,
    calc
}


% --- Document Settings ---
\onehalfspacing
\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=black}

% --- Title Page ---
\begin{document}

\begin{titlepage}
    \centering
    \vspace*{1cm}
    {\Large \textbf{Research Project - Intermediate Report}}\\[1.5cm]

    {\huge \textbf{District-Level Dengue Outbreak Early Warning System for Sri Lanka}}\\[3cm]

    \textbf{Group Number: Group 10}\\[4.5cm]

    {\textbf{Group Members:}}\\[0.3cm]
    \begin{table}[h]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Name}      & \textbf{Student ID} \\ \hline
            Kavindu Amalka     & SE/2021/008         \\ \hline
            Hiruni Imasha      & SE/2021/053         \\ \hline
            Sandaru Mihiran    & SE/2021/010         \\ \hline
            Sachini Weerakkody & SE/2021/018         \\ \hline
        \end{tabular}
    \end{table}

    \vfill
    Date: \today
\end{titlepage}

\tableofcontents
\newpage

% --- 1. INTRODUCTION & PROJECT OVERVIEW ---
\section{Introduction \& Project Overview}

Dengue fever is a critical, hyperendemic public health crisis in Sri Lanka. Its transmission is closely linked to seasonal monsoonal climate shifts, yet current national management strategies remain reactive, relying on retrospective hospital records rather than anticipation. Consequently, vector control measures are often deployed after outbreaks peak, causing sub-optimal medical resource allocation.  

To transition from crisis management to proactive strategic prevention, this research introduces an early warning system covering all 26 Regional Director of Health Services (RDHS) areas across Sri Lanka. By leveraging machine learning architectures, the system integrates historical epidemiological trends with micro-climatic indicators to deliver a 4-week-ahead outbreak forecasting index. This early warning buffer empowers Public Health Inspectors (PHIs) to precision-target vector populations and mobilize emergency medical resources before clinical surges occur.  
\subsection{Research Objectives}
\begin{itemize}
    \item \textbf{Data Fusion:} Integrate historical weekly dengue cases extracted from the Epidemiology Unit’s Weekly Epidemiological Reports with synchronized meteorological data (Rainfall, Temperature, Humidity) from the NASA POWER API into a unified 26 RDHS areas longitudinal panel dataset.
    \item \textbf{Dual-Stream Modeling:} Implement parallel LSTM and XGBoost machine learning pipelines to execute Regression (case forecasting) and Classification (outbreak risk mapping) within a single execution block.
    \item \textbf{Regional Adaptation \& Lag Analysis:} Isolate RDHS-specific climate-to-outbreak time lags to capture the unique epidemiological dynamics between Sri Lanka's Wet and Dry zones.
    \item \textbf{National Scale Generalization:} Expand predictive modeling across all 26 RDHS health administration regions simultaneously, overcoming the geographical limitations of prior localized studies.
    \item \textbf{Decision-Support Interface:} Build an interactive React.js web dashboard that translates complex machine learning forecasts into intuitive geospatial risk visualizations for health authorities.
\end{itemize}


\clearpage
% --- 2. COMPREHENSIVE LITERATURE REVIEW ---
\section{Comprehensive Literature Review}
\label{sec:lit_review}

\subsection{Systematic Review of Relevant Sources}
% TODO: Add systematic review of minimum 20 relevant sources.
This subsection will present a systematic review of at least 20 relevant research papers, white papers, and articles concerning dengue forecasting, meteorological influence, human mobility modeling, and machine learning architectures.

\subsection{Critical Analysis of Existing Work}
% TODO: Compare and critically analyze previous methodologies (ARIMA, LSTM, XGBoost, etc.).
A critical evaluation of existing studies will be detailed here, highlighting their strengths, weaknesses, geographical biases (e.g., Western Province bias), and gaps in translating model output into active systems.

\subsection{Clear Positioning of Your Research}
% TODO: Explain where this research fits and how it improves on existing methods.
This section positions our research as a dual-stream forecasting and classification framework that incorporates environmental, mobility, and larval data specifically calibrated for the distinct wet/dry zones of Sri Lanka.

\subsection{Journal/Conference Publication Plan}
% TODO: Document plan for submitting a standalone literature review paper.
Consideration of compiling this systematic review into a standalone literature review paper to be submitted to a relevant peer-reviewed journal or international conference.


\clearpage
% --- 3. SOLID EXPERIMENTAL SETUP / RESEARCH DESIGN ---
\section{Experimental Setup \& Research Design}
\label{sec:experimental_setup}

\subsection{Detailed Methodology Description}

This research follows a quantitative experimental research design combined with a software prototype development approach. The main objective is to develop a district/RDHS-level dengue early warning framework for Sri Lanka using weekly dengue case data and NASA POWER weather data. The system is designed to predict dengue risk four weeks ahead, so that public-health officers can identify possible outbreak situations before they become severe.

The first phase of the methodology focuses on data collection from two primary sources. Weekly district/RDHS-level dengue case data is obtained from the Weekly Epidemiological Reports published by the Epidemiology Unit, Ministry of Health, Sri Lanka. This dataset provides historical dengue incidence counts, which are used as the target variable in the study. The second source is meteorological data retrieved from the NASA POWER API using representative geographic coordinates for all 25 districts of Sri Lanka, along with Kalmunai as an additional administrative unit. The key weather variables considered include rainfall, temperature, and relative humidity. To automate data retrieval, a Python-based web application was developed using the Streamlit framework, enabling users to select a custom date range and extract weather data for all districts simultaneously. Since NASA POWER provides daily-resolution data, the system aggregates these values into weekly summaries using the Pandas library, ensuring compatibility with epidemiological reporting. The resulting weather data is then aligned with dengue case records based on district/RDHS area and epidemiological week.

The second phase involves data preprocessing and integration to construct a unified analytical dataset. Weather data is standardized by formatting district names, calendar years, dates, and epidemiological week identifiers to ensure consistency across all records. Daily meteorological observations are converted into weekly aggregates, where rainfall is represented as total weekly rainfall, while temperature and humidity are computed as weekly average values. The dengue and weather datasets are then merged using common fields such as district/RDHS area, year, and epidemiological week. The final output is a structured district/RDHS-week panel dataset, where each record represents a specific geographic area in a given epidemiological week, containing both dengue incidence and corresponding weather conditions.

The third phase of the study focuses on exploratory data analysis (EDA) and climate-lag analysis. Climate-lag analysis is used to identify the delayed effect of weather conditions on dengue transmission, since changes in rainfall, temperature, and humidity do not result in immediate increases in dengue cases. Instead, there is a biological and epidemiological delay due to mosquito breeding cycles, vector development, and reporting time. To capture this relationship, different lag periods ranging from 1-week to 12-weeks are tested to determine which weather conditions show the strongest delayed association with dengue incidence across districts and, where applicable, across climate zones such as wet, dry, and intermediate zones. In addition, exploratory analysis is performed using Power BI to visualize temporal and spatial patterns in the dataset. Time-series visualizations, such as the weekly dengue trend chart, are generated to observe outbreak peaks, seasonal fluctuations, and long-term variations in dengue incidence, supporting the identification of epidemic periods and potential weather–disease relationships.

The fourth stage is feature engineering. Feature engineering means creating additional model input variables from the original dengue and weather data. These features help the models learn delayed and seasonal dengue patterns. Dengue history features may include previous week cases, two-week lag cases, four-week lag cases, and rolling average case counts. Weather features may include rainfall lag, temperature lag, humidity lag, and rolling weather averages. A rolling average means the average value over recent weeks, such as the average dengue cases during the previous four weeks. Seasonal features such as epidemiological week, month, Southwest monsoon, Northeast monsoon, and inter-monsoon periods will also be added. These features help the models learn both short-term and seasonal dengue behaviour.

The fifth stage is comparative modelling. This research compares three representative modelling approaches: SARIMAX, LightGBM, and LSTM. SARIMAX is selected as the main statistical baseline because it can model time-series seasonality and external weather variables, and it has shown strong performance in Sri Lankan dengue forecasting literature. LightGBM is selected as the machine learning model because it can handle nonlinear relationships in tabular data, such as dengue lags, weather lags, monsoon indicators, and district-level features. LSTM is selected as the deep learning model because it can learn sequential patterns from previous weeks of dengue and weather data.

The modelling output follows a dual-stream prediction design. The first stream is a regression stream, where the model predicts the expected number of dengue cases four weeks ahead. The second stream is a classification stream, where the system predicts whether a district/RDHS area is likely to enter an outbreak or high-risk condition four weeks ahead. For classification, a district-specific alert threshold can be created using historical dengue case patterns, such as historical mean plus two standard deviations. This allows the system to provide both a numerical forecast and a practical warning decision.

After model training, the models will be evaluated using regression and classification metrics to identify the most suitable model for dashboard integration. This phase focuses on comparing different models based on their predictive performance and selecting the best-performing model for deployment.

The final stage is dashboard prototype development. The dashboard will mainly display predicted dengue case counts, alert status, dengue trends, and weather trends. If implementation time allows, district risk maps and main contributing weather-lag factors will also be included. This makes the research useful not only as a prediction experiment but also as a public-health decision-support system.

\subsection{Tools, Technologies, and Frameworks Selected}

The tools and technologies selected for this research are chosen based on data processing, modelling, evaluation, explainability, and dashboard development requirements.

\begin{table}[htbp]
\centering
\caption{Selected Tools, Technologies, and Frameworks}
\label{tab:tools_technologies}
\small
\begin{tabular}{p{3.4cm} p{4.2cm} p{6.4cm}}
\toprule
\textbf{Category} & \textbf{Selected Tools / Technologies} & \textbf{Purpose} \\
\midrule
Programming language & Python & Main language for data processing, modelling, and evaluation. \\

Data handling & Pandas, Power BI, NumPy & Preparing the weather dataset, integrating dengue case values, generating features, and analysing dengue-weather patterns. \\

Weather data source & NASA POWER API & Collecting rainfall, temperature, and humidity values using representative district/RDHS coordinates. \\

Statistical modelling & statsmodels / pmdarima & Developing and testing SARIMAX time-series models. \\

Machine learning & LightGBM, scikit-learn & Training LightGBM models and supporting preprocessing, validation, and metric calculation. \\

Deep learning & TensorFlow/Keras or PyTorch & Developing LSTM-based time-series models to learn sequential dengue-weather patterns. \\

Visualization & Matplotlib, Plotly & Creating dengue trend graphs, weather graphs, lag-analysis plots, model result visualizations, and interactive dashboard charts. \\

Explainability & SHAP / Feature importance & Explaining which features contribute most to dengue predictions, especially for the LightGBM model. \\

Model evaluation & scikit-learn metrics, NumPy, Pandas & Calculating MAE, RMSE, MAPE, Precision, Recall, F1-score, and ROC-AUC. \\

Development & Streamlit, React & Building the early warning dashboard prototype. Streamlit is suitable for fast Python-based dashboard development, while React may be considered later for a more customized frontend. \\

Version control & GitHub & Managing source code, documentation, experiment versions, and collaboration. \\
\bottomrule
\end{tabular}
\end{table}





The study uses Python as the primary programming language due to its strong support for data processing, time-series modelling, machine learning, and deep learning. Data handling is performed using Pandas, NumPy, and Power BI to prepare the weather dataset, integrate dengue case data, generate features, and analyse dengue–weather relationships. NASA POWER API is used as the weather data source to collect rainfall, temperature, and humidity values based on district/RDHS coordinates.

Statistical modelling is carried out using statsmodels or pmdarima for SARIMAX-based time-series forecasting, while LightGBM and scikit-learn are used for machine learning model development, preprocessing, validation, and performance evaluation. Deep learning models are implemented using TensorFlow/Keras or PyTorch to capture sequential dengue–weather patterns.

For visualization, Matplotlib and Plotly are used to generate dengue trends, weather patterns, lag-analysis results, and model outputs. Model interpretability is achieved using SHAP and feature importance techniques, particularly for the LightGBM model, to identify key influencing variables.

Model evaluation is performed using scikit-learn metrics along with NumPy and Pandas to compute performance measures such as MAE, RMSE, MAPE, Precision, Recall, F1-score, and ROC-AUC.

For system development, Streamlit and React are used to build the early warning dashboard prototype, where Streamlit has already been implemented for weather data handling and will be extended for dashboard development. GitHub is used for version control, supporting code management, documentation, and collaboration throughout the project.

\subsection{Research Framework}

The proposed research framework contains seven main stages: data acquisition, data preprocessing and integration, exploratory data analysis and climate-lag analysis, feature engineering, comparative modelling, validation and evaluation, and dashboard prototype development.

\begin{figure}[htbp]
\centering
\resizebox{0.95\textwidth}{!}{
\begin{tikzpicture}[
    node distance=1.15cm,
    every node/.style={font=\small},
    input/.style={rectangle, rounded corners, draw=blue!70, fill=blue!10, thick, align=center, minimum width=4.6cm, minimum height=1cm},
    process/.style={rectangle, rounded corners, draw=gray!80, fill=gray!10, thick, align=center, minimum width=7.2cm, minimum height=1.1cm},
    model/.style={rectangle, rounded corners, draw=orange!80, fill=orange!15, thick, align=center, minimum width=3.4cm, minimum height=1cm},
    output/.style={rectangle, rounded corners, draw=green!70!black, fill=green!10, thick, align=center, minimum width=4.2cm, minimum height=1cm},
    arrow/.style={-Latex, thick}
]

\node[input] (dengue) {Weekly Dengue Case Data\\Epidemiology Unit / WER};
\node[input, right=1.4cm of dengue] (weather) {NASA POWER Weather Data\\Rainfall, Temperature, Humidity};

\node[process, below=1.2cm of $(dengue)!0.5!(weather)$] (prep) {Data Preprocessing \& Integration\\District/RDHS + Year + Epidemiological Week};
\node[process, below=of prep] (eda) {Exploratory Data Analysis\\District trends, weather trends, wet/dry-zone patterns};
\node[process, below=of eda] (lag) {Climate-Lag Analysis\\Testing 1 to 12 week weather lags};
\node[process, below=of lag] (feature) {Feature Engineering\\Dengue lags, weather lags, rolling averages, monsoon and zone features};

\node[model, below left=1.2cm and 2.6cm of feature] (sarimax) {SARIMAX\\Statistical Baseline};
\node[model, below=1.2cm of feature] (lgbm) {LightGBM\\Machine Learning Model};
\node[model, below right=1.2cm and 2.6cm of feature] (lstm) {LSTM\\Deep Learning Model};

\node[process, below=1.5cm of lgbm] (dual) {Dual Prediction Output\\Regression: 4-week-ahead case count\\Classification: Alert / No Alert};
\node[process, below=of dual] (validation) {Validation \& Evaluation\\MAE, RMSE, MAPE, Precision, Recall, F1-score, ROC-AUC};
\node[output, below=of validation] (dashboard) {Early Warning Dashboard Prototype\\Predicted cases, alert level, trends, optional risk map};

\draw[arrow] (dengue) -- (prep);
\draw[arrow] (weather) -- (prep);
\draw[arrow] (prep) -- (eda);
\draw[arrow] (eda) -- (lag);
\draw[arrow] (lag) -- (feature);
\draw[arrow] (feature) -- (sarimax);
\draw[arrow] (feature) -- (lgbm);
\draw[arrow] (feature) -- (lstm);
\draw[arrow] (sarimax) -- (dual);
\draw[arrow] (lgbm) -- (dual);
\draw[arrow] (lstm) -- (dual);
\draw[arrow] (dual) -- (validation);
\draw[arrow] (validation) -- (dashboard);

\end{tikzpicture}
}
\caption{Proposed dengue early warning research framework}
\label{fig:research_framework}
\end{figure}

The framework is designed to support four-week-ahead dengue early warning. The input layer contains weekly dengue case data and NASA POWER weather data. The preprocessing layer converts all data into a common district/RDHS-week format. The analysis layer studies dengue trends, weather patterns, wet-zone and dry-zone differences, and delayed climate effects. The feature engineering layer creates model-ready variables such as dengue lags, weather lags, rolling averages, monsoon indicators, and climate-zone indicators.

The modelling layer compares SARIMAX, LightGBM, and LSTM under the same prediction setup. SARIMAX acts as the statistical baseline model. LightGBM acts as the main machine learning model for structured lag-based features. LSTM acts as the deep learning sequence model. This comparison allows the research to evaluate whether statistical, machine learning, or deep learning approaches are more suitable for Sri Lankan district/RDHS-level dengue early warning using NASA POWER weather data.

The output layer produces both predicted case counts and outbreak alerts. The validation and evaluation layer checks whether the models are reliable before integrating the selected model into the dashboard. Finally, the dashboard layer presents the results in a practical format for decision-making.

\subsection{Validation Approach}

The validation approach is designed to test the models under realistic forecasting conditions. Since dengue forecasting is a time-series problem, random train-test splitting is not suitable. Random splitting can mix past and future data, which may give unrealistic performance. Therefore, this research will use walk-forward validation.

In walk-forward validation, the model is trained using past data and tested on future data. This better represents a real early warning system because the model must always predict future dengue risk using only information available up to the current week.

An example validation setup is shown below:

\begin{center}
\begin{tabular}{p{5cm} p{5cm}}
\toprule
\textbf{Training Period} & \textbf{Testing Period} \\
\midrule
2016--2020 & 2021 \\
2016--2021 & 2022 \\
2016--2022 & 2023 \\
2016--2023 & 2024 \\
\bottomrule
\end{tabular}
\end{center}

All models will be evaluated using the same dataset, same prediction horizon, and same validation method. This ensures a fair comparison between SARIMAX, LightGBM, and LSTM. Before validation, the dataset will be checked to ensure that dengue cases are correctly aligned with the same district/RDHS area and epidemiological week as the weather variables.

For the regression stream, the models will be evaluated using the metrics shown in Table~\ref{tab:regression_metrics}.

\begin{table}[htbp]
\centering
\caption{Regression Evaluation Metrics}
\label{tab:regression_metrics}
\begin{tabular}{p{3.5cm} p{9cm}}
\toprule
\textbf{Metric} & \textbf{Purpose} \\
\midrule
MAE & Measures the average size of prediction errors. \\

RMSE & Penalizes larger prediction errors more strongly. \\

MAPE & Shows prediction error as a percentage. \\

Forecast bias & Shows whether the model usually overpredicts or underpredicts. \\
\bottomrule
\end{tabular}
\end{table}

For the classification stream, the models will be evaluated using the metrics shown in Table~\ref{tab:classification_metrics}.

\begin{table}[htbp]
\centering
\caption{Classification Evaluation Metrics}
\label{tab:classification_metrics}
\begin{tabular}{p{3.5cm} p{9cm}}
\toprule
\textbf{Metric} & \textbf{Purpose} \\
\midrule
Precision & Measures how many predicted alerts were actually correct. \\

Recall & Measures how many real outbreak situations were detected. \\

F1-score & Balances precision and recall. \\

ROC-AUC & Measures the model's ability to separate alert and non-alert situations. \\
\bottomrule
\end{tabular}
\end{table}

In outbreak prediction, recall is especially important because missing a real outbreak can be more harmful than issuing a false warning. Therefore, the final model selection will consider not only overall accuracy but also the model's ability to detect high-risk dengue situations early.

The final selected model will be chosen based on both prediction performance and practical usefulness. A model with slightly lower numerical accuracy but better outbreak detection and interpretability may be more useful for a public-health early warning system. Therefore, the validation process will consider regression accuracy, classification performance, alert reliability, and interpretability before selecting the final model for dashboard integration.









\clearpage
% --- 4. DATA COLLECTION PROGRESS ---
\section{Data Collection Progress}
\label{sec:data_collection}

\subsection{Data Sources Identified and Accessed}
% TODO: Detail the institutional sources (Epidemiology Unit, Met Department, Transit authorities).
Table~\ref{tab:data_sources} summarizes the data sources identified, accessed, and integrated into our data engineering pipeline.

\begin{table}[h]
    \centering
    \caption{Identified and Accessed Data Sources}
    \label{tab:data_sources}
    \begin{tabular}{lllp{5cm}}
        \toprule
        \textbf{Dataset} & \textbf{Source Institution} & \textbf{Temporal Resolution} & \textbf{Access Status} \\
        \midrule
        Epidemiological Cases & Epidemiology Unit & Weekly & Accessed \& Downloaded (2017--2026) \\
        Meteorological Factors & Dept. of Meteorology & Weekly / Daily & Accessed (Rainfall, Temp, Humidity) \\
        Commuter Flows & Transit Authorities & Static Network & Under Negotiation / Public Schedules \\
        Larval Indices & PHI Reports & Monthly / Weekly & Under Request \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Data Collection Methods Implemented}
% TODO: Describe scripts, APIs, or scraping used to gather the data.
Description of automated data retrieval scripts, manual collation processes, and pipeline steps implemented to download and store datasets.

\subsection{Sample Data Collected and Documented}
% TODO: Show statistical summaries or structures of the collected sample data.
Sample data snippets, feature formats, schemas, and basic descriptive statistics will be documented in this section.

\subsection{Challenges in Data Acquisition}
% TODO: Document challenges like missing data, administrative approvals, or API limitations.
A discussion on data gaps, delays in administrative permissions, format inconsistencies, and the respective imputation/alignment methods used to address them.


\clearpage
% --- 5. PRELIMINARY RESULTS ---
\section{Preliminary Results}
\label{sec:preliminary_results}

\subsection{Initial Findings or Prototype Functionality}
% TODO: Describe the working baseline model or early dashboard prototype.
Details regarding the baseline regression and classification models, as well as mockups/features of the early React.js dashboard layout.

\subsection{Preliminary Analysis or Testing Results}
% TODO: Share initial metric scores (e.g. baseline MAE, accuracy) of the models.
Initial model performance values from training on historical Gampaha/Colombo district data.

\subsection{Visualizations, Metrics, or Demonstrations}
% TODO: Add preliminary charts showing case predictions vs actual historical trends.
Graphical representations of case trends, feature correlations, and early spatial hazard maps.

\subsection{Discussion of Early Insights}
% TODO: Discuss what the early findings imply about predictability, weather lags, or mobility features.
Insights drawn from early model outputs, such as the significance of 4--8 week temperature/rainfall lags and inter-district commuter risk indicators.

\newpage
% --- REFERENCES ---
\nocite{*}
\printbibliography

\end{document}
